---
title: "Project_Milestone4"
author: "Liza Lutker and Emily Liu"
format: html
editor: visual
---

## Project Milestone 4

Instructions:

Please submit an Rmd or Qmd and publish an html file on RPubs with the following:

1.  Final datasets for creation of visualization

    -   Join all datasets together - join in asthma data.

    -   Calculate any remaining data elements needed for analysis - not necessary to create, but should drop all race and ethnicity groups except white, Black, and Hispanic (due to NA variables) and overall group.

    -   Show code used to create joined dataset, but please do not print full data frame output (showing data structure with `str()` is okay)

2.  Visualizations (at least one per group member)

3.  Your task is to utilize these datasets to create county-level measures to accomplish two goals. The first is to compare county asthma ED rates with a county CES measure to assess if there appears to be a correlation. The second is to compare asthma ED rates with county-level summaries for specific environmental measures to determine if those specific measures may be worth further investigation.

    -   One print quality table as requested in scenario

        -   Table 1 with all relevant measures / overall, stratify by CES measure

# Joining Data

Code to join data sets and provide appropriate labels for table output.

```{r, echo = FALSE}
# Install/load packages
library(tidyverse)
install.packages("gtsummary")
library(gtsummary)
install.packages("labelled")
library(labelled)

# Load in cleaned datasets
ed_summary <- 
  readRDS('./Project_Data_Sources/scenario_3/ed_summary.RDS')
ces_measures_summary <- 
  readRDS('./Project_Data_Sources/scenario_3/ces_measures_summary.RDS')
ces_score_summary <- 
  readRDS('./Project_Data_Sources/scenario_3/ces_score_summary.RDS')
names(ed_summary)

# Join datasets using variable "county_name"
# NOTES:(1) Drops small racial and ethnic groups with suppressed cells. Keep White, Black, and Hispanic, and overall ED asthma rates. (2) Add labels to table variables. 

joined_data <- ces_score_summary %>% 
  left_join(., ces_measures_summary, by = "county_name") %>%
  left_join(., ed_summary, by = "county_name") %>%
  select(-c(api, aian, nhpi, multi)) %>% 
  set_variable_labels(
  median_ces_perc_county = "Median CES score percentile", 
  median_ces_abv_perc_50th = "Median CES score, 50th percentile and above", 
  pm_25_mean = "Mean PM2.5",
  traffic_median = "Traffic median",
  poverty_perc_mean = "Mean poverty",
  poverty_abv_75th = "Mean poverty above 75th percentile",
  total = "Total age-adjusted rate of ED asthma visits",
  white = "Age-adjusted rate of ED asthma visits - White patients",
  black = "Age-adjusted rate of ED asthma visits - Black patients",
  hisp = "Age-adjusted rate of ED asthma visits - Hispanic patients"
)

str(joined_data)
```

# Table 1

Code for output table.

```{r}
#Table Output
joined_data %>% 
  select(-c(county_name, median_ces_abv_perc_75th, median_ces_blw_perc_25th)) %>% 
  tbl_summary(by = median_ces_abv_perc_50th,
              missing_text = "Rate suppressed for small cell count") %>% 
  add_overall() %>%
  add_n() %>%
  modify_header(label ~ "**Variable**") %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ 
  "**Median CES score above 50th percentile**") %>%
  modify_footnote(
    all_stat_cols() ~ "Median (IQR) or Frequency (%)"
  ) %>%
  modify_caption("**Table 1. County Characteristics**") %>%
  bold_labels()
```

**TABLE INTERPRETATION:** Counties with CES median scores above the 50th percentile appear to have higher mean PM 2.5, higher median traffic, and more poverty. Additionally, these counties have age-adjusted rates of ED asthma visits total and stratified by racial and ethnic group.

# Plots

-   One print quality plot or chart as requested in scenario

    -   \*Maybe do this for all measures (CES measure + specific environmental measures).

    -   CES measure (1/0) - x axis; boxplot for asthma rates among Black individuals

    -   CES measure (1/0) - x axis; boxplot for asthma rates among white individuals

    -   CES measure (1/0) - x axis; boxplot for asthma rates among Hispanic individuals

    -   CES measure (1/0) - x axis; boxplot for asthma rates overall

-   For groups of 3, one additional print quality table or plot of your choice (can support the requested data in the scenario, or answer a different question using the same data sources)

1.  For each visual, include

    -   Code used to generate visual

    -   Legend (if necessary)

    -   1-2 sentence interpretation

    -   NOTE:

        -   Please make sure the visual can stand-alone, meaning it includes enough information in title, legend, and footnote so that a person who sees only the visualization could understand what is being presented.

        -   Please also make sure column names, axis labels, and any other labels are meaningful and not just the name of the variable (ex: "County" rather than "county_name")

2.  Html file that is professionally prepared for presentation and published to RPubs

    -   Only the necessary information is in the output (e.g., suppress entire data frame outputs but showing data structure with `str()` is okay)

    -   Show your work by "echoing" code used in Rmd/Qmd file to create your tables and graphs. \`\`\`{r, echo = TRUE} in code chunk preferences.

    -   Use of headers and sub headers to create an organized document

    -   For this milestone please make sure code for visualizations are included in the final file
