---
title: "Project Milestone 3"
author: "Liza Lutker and Emily Liu"
format:
  html:
    code-overflow: wrap
editor: visual
---

# Data Cleaning

**Load in data**

```{r}
#| output: false
#| warning: false

library(tidyverse)
library(janitor)

#bring in files and clean up column names
ed <- read.csv('./Project_Data_Sources/scenario_3/chhs_asthma_ed.csv') %>% 
  clean_names()

ces_scores <- read.csv('./Project_Data_Sources/scenario_3/calenviroscreen_scores_demog_2021.csv') %>% clean_names()

ces_measures <- read.csv('./Project_Data_Sources/scenario_3/calenviroscreen_measures_2021.csv') %>% clean_names()
```

**Data Cleaning**

***Part 1: 2021 Cal Enviroscreen Scores***

-   Change "county" variable to "county-name" for consistency between datasets

-   Created 4 new variables by county

    -   *median_ces_perc_county* = median CES percentile score by county (type: num)

    -   *median_ces_abv_perc_75th* = flag that indicates if the county median CES percentile score is equal to or **above** the 75th percentile (type: factor)

    -   *median_ces_abv_perc_50th* = flag that indicates if the county median CES percentile score is equal to or **above** the 50th percentile (type: factor)

    -   *median_ces_blw_perc_25th* = flag that indicates if the county median CES percentile score is equal to or **below** the 25th percentile (type: factor)

-   Output: Dataset named **ces_score_summary** that can be joined by *county_name*

```{r}
#| output: false
# ces_scores %>% select(county) %>% distinct() #examine unique county names
# length(unique(ces_scores$county)) #58

# Change county name from "Alameda County" -> "ALAMEDA"
ces_score_updated <- ces_scores %>% 
  mutate(county_name = str_to_upper(str_trim(str_remove(county, "County"))))

# ces_score_updated %>% select(county_name) %>% distinct() #check new county name

ces_score_summary <- ces_score_updated %>% group_by(county_name) %>%
  summarize(median_ces_perc_county = median(ces_4_0_percentile, na.rm = T),
            median_ces_abv_perc_75th = if_else(median_ces_perc_county >= 75, 1, 0),                        median_ces_abv_perc_50th = if_else(median_ces_perc_county >= 50, 1, 0),
            median_ces_blw_perc_25th = if_else(median_ces_perc_county <= 25, 1, 0))

ces_score_summary$median_ces_abv_perc_75th <- 
  factor(ces_score_summary$median_ces_abv_perc_75th, 
         levels = c(0, 1), 
         labels = c("No", "Yes"))

ces_score_summary$median_ces_abv_perc_50th <- 
  factor(ces_score_summary$median_ces_abv_perc_50th, 
         levels = c(0, 1), 
         labels = c("No", "Yes"))

ces_score_summary$median_ces_blw_perc_25th <- 
  factor(ces_score_summary$median_ces_blw_perc_25th, 
         levels = c(0, 1), 
         labels = c("No", "Yes"))

# str(ces_score_summary)
```

***Part 2: 2021 Cal Enviroscreen Measures***

-   Change "county" variable to "county-name" for consistency between datasets

-   Created 4 new variables by county

    -   *pm_25_mean* = county mean PM 2.5 value (type: num)

    -   *traffic_median*= county median traffic value (type: num)

    -   *poverty_perc_mean* = county mean of poverty percentile (type: num)

    -   *poverty_abv_75th* = flag that indicates if the county mean poverty percentile is equal to or above the 75th percentile. (type: factor)

-   Output: Dataset named **ces_measures_summary** that can be joined by *county_name*

```{r}
#| output: false
# ces_measures %>% select(california_county) %>% distinct()
# length(unique(ces_measures$california_county)) #58

# Change county name from "Alameda" -> "ALAMEDA"
ces_measures_updated <- ces_measures %>% 
  mutate(county_name = str_to_upper(str_trim(california_county)))
# ces %>% select(county_name) %>% distinct()

ces_measures_summary <- ces_measures_updated %>% group_by(county_name) %>%
  summarize(pm_25_mean = mean(pm2_5, na.rm = TRUE),
            traffic_median = median(traffic, na.rm = TRUE), 
            poverty_perc_mean = mean(poverty_pctl, na.rm = TRUE), 
            poverty_abv_75th = if_else(poverty_perc_mean >= 75, 1, 0))

ces_measures_summary$poverty_abv_75th <- 
  factor(ces_measures_summary$poverty_abv_75th, 
         levels = c(0, 1), 
         labels = c("No", "Yes"))
```

***Part 3: Asthma Rates***

-   Cleaned up category names for age.

-   Change "california_county" variable to "county_name" for consistency between datasets

    -   Dropped values with county_name = "CALIFORNIA" since that is an overall total among all counties in California.

-   Pivot table so we have one row per county (58 rows)

-   Output: Dataset named **ed_summary** that can be joined by *county_name*

-   **NOTE:**

    -   There are discrepancies of racial and ethnic categorizations of Asian, API, and NHPI between years. See notes in the code comments.

    -   There are many NA values for the race-stratified data due to "Count and rate suppressed in accordance with data de-identification guidelines".

    -   For age-adjusted rate, data dictionary indicates that rates based on counts less than 12 are suppressed due to statistical instability.

    -   **Needed data decisions:**

        -   How do we want to handle these data in the analysis phase?

        -   Are there specific racial and ethnic groups that we want to drop from the analysis due to small cell counts and discrepancies in categorization?

```{r}
#| output: false

#look at values for variables and see if they need any adjustment

# summary(ed) # NOTE: will want to factor year
# unique(ed$strata)
# unique(ed$strata_name)
# table(ed$strata_name, ed$year)

# NOTE: these values are reading in oddly and need fixing plus there are different ways racial groups are ID'ed across years (Asian only used in 2019, otherwise use Asian/PI; Multi-race used in 2020 but otherwise use Multi-Race. Not sure what to do about the fact that Asian/PI and NHPI seem to be used concurrently in 2019 and 2020 - leave as is) 

# unique(ed$age_group)
# NOTE: age is also reading in oddly and need recoding

ed_1 <- ed %>% 
  mutate(str_name = case_when(
      strata_name == "All ages" ~ "all_ages",
      strata_name == "0\x9617 years" ~ "0_17",
      strata_name == "18+ years" ~ "18plus",      
      strata_name == "0\x964 years" ~ "0_4",
      strata_name == "5\x9617 years" ~ "5_17",
      strata_name == "18\x9664 years" ~ "18_64",
      strata_name == "65+ years" ~ "65plus",
      strata_name == "White" ~ "WHITE",
      strata_name == "Black" ~ "BLACK",
      strata_name == "Hispanic" ~ "HISP",
      strata_name %in% c("Asian/PI", "Asian") ~ "API",
      strata_name == "NHPI" ~ "NHPI",
      strata_name == "AI/AN" ~ "AIAN",
      strata_name %in% c("Multi-Race", "Multi-race") ~ "MULTI",
      )) %>% 
  mutate(age_grp = case_when(
      age_group == "All ages" ~ "all_ages",
      age_group == "0\x9617 years" ~ "0_17",
      age_group == "18+ years" ~ "18plus",      
      age_group == "0\x964 years" ~ "0_4",
      age_group == "5\x9617 years" ~ "5_17",
      age_group == "18\x9664 years" ~ "18_64",
      age_group == "65+ years" ~ "65plus",
  )) %>% 
  mutate(yr=as.factor(year)) %>% 
  select(!c(strata_name, age_group, year))

# Check updated values
# unique(ed1$str_name)
# table(ed1$strata_name, ed1$str_name, useNA = "always")
# unique(ed1$age_grp)
# table(ed1$age_group, ed1$age_grp, useNA = "always")
# unique(ed1$yr)

# ed_1 %>% select(county) %>% distinct()
# length(unique(ed$county)) #59

# drop california county (overall rows)
ed_1 <- ed_1 %>% filter(county != "CALIFORNIA") %>% rename(county_name = county)

# ed %>% select(county_name) %>% distinct()

# subset to years, strata (all age groups by race) of interest and pivot to have one row per county (since examining all ages combined, use age-adjusted values)

ed_summary <- ed_1 %>% 
  filter(yr == "2020") %>% 
  filter(age_grp == "all_ages") %>% 
  select(!c(strata, number_of_ed_visits, age_grp, yr)) %>% 
  pivot_wider(names_from = str_name, values_from = age_adjusted_ed_visit_rate) %>% 
  rename(TOTAL = all_ages) %>% clean_names()

#names(ed_summary)
```

# Data Dictionary

***Part 1: 2021 Cal Enviroscreen Scores***

-   **Dataset:** ces_score_summary
-   **Variables:**
    -   *county_name* = name of California county in all caps format (e.g. "ALAMEDA") (type: character)

    -   *median_ces_perc_county* = median CES percentile score by county (type: num)

    -   *median_ces_abv_perc_75th* = flag that indicates if the county median CES percentile score is equal to or **above** the 75th percentile (type: factor)

    -   *median_ces_abv_perc_50th* = flag that indicates if the county median CES percentile score is equal to or **above** the 50th percentile (type: factor)

    -   *median_ces_blw_perc_25th* = flag that indicates if the county median CES percentile score is equal to or **below** the 25th percentile (type: factor)

***Part 2: 2021 Cal Enviroscreen Measures***

-   **Dataset**: ces_measures_summary

-   **Variables**

    -   *county_name* = name of California county in all caps format (e.g. "ALAMEDA") (type: character)

    -   *traffic_median*= county median traffic value (type: num)

    -   *poverty_perc_mean* = county mean of poverty percentile (type: num)

    -   *poverty_abv_75th* = flag that indicates if the county mean poverty percentile is equal to or above the 75th percentile. (type: factor)

***Part 3: Asthma Rates***

-   **Dataset:** ed_summary

-   **Variables:**

    -   *county_name* = name of California county in all caps format (e.g. "ALAMEDA") (type: character)

    -   *total* = age-adjusted rate of ED visits for all racial and ethnic groups. (type = num)

    -   *white* = age-adjusted rate of ED visits for racial and ethnic group of White (type = num)

    -   *black* = age-adjusted rate of ED visits for racial and ethnic group of Black (type = num)

    -   *hisp* = age-adjusted rate of ED visits for racial and ethnic group of Hispanic (type = num)

    -   *api* = age-adjusted rate of ED visits for racial and ethnic group of Asian and Pacific Islander (type = num)

    -   *aian* = age-adjusted rate of ED visits for racial and ethnic group of Alaskan Native and American Indian (type = num)

    -   *nhpi* = age-adjusted rate of ED visits for racial and ethnic group of Native Hawaiian and Pacific Islander (type = num)

    -   *multi* = age-adjusted rate of ED visits for racial and ethnic group of multiracial (type = num)

        # Frequencies

```{r}
# install.packages("gtsummary")
library(gtsummary)

# join ces_score_summary and ces_measures_summary by county name
freq_data <- ces_score_summary %>% left_join(., ces_measures_summary, by = "county_name") 

# 
# frequencies 
freq_data %>% 
  select(median_ces_perc_county, median_ces_abv_perc_75th, 
         median_ces_abv_perc_50th, median_ces_blw_perc_25th,
         traffic_median, poverty_perc_mean, poverty_abv_75th) %>% 
  tbl_summary(
    label = c(median_ces_perc_county ~ "County Median CES Percentile",
              median_ces_abv_perc_75th ~ "County Median CES Percentile above 75",
              median_ces_abv_perc_50th ~ "County Median CES Percentile above 50", 
              median_ces_blw_perc_25th ~ "County Median CES Percentile below 25",
              traffic_median ~ "County Median Traffic", 
              poverty_perc_mean ~ "County Mean Poverty Percentile", 
              poverty_abv_75th ~ "County Mean Poverty Percentile above 75")
  )

```
