---
title: "Project Milestone 2"
author: "Emily Liu & Liza Lutzker"
format: html
editor: visual
---

## Description of the datasets

### What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)

-   Dataset 1 (calenviroscreen_measures_2021.csv) contains multiple specific environmental measures and Dataset 2 (calenviroscreen_scores_demog_2021.csv) contains overall CalEnviroScreen (CES) scores and demographics. Both are from CalEnviroScreen, provide data at the census-tract level (accompanied by county of tract location), and use 2019 ACS estimates for demographics. 

-   Dataset 3 (chhs_asthma_ed.csv) is from CA Health and Human Services Dept and has annual age-adjusted asthma ED visits/rates by county from 2015 through 2020. 

### How does the dataset relate to the group problem statement and question?

Dataset 3 provides outcome data (acute asthma as defined by asthma ED visit). Dataset 1 provides the exposure data related to the "specific environmental measures" to be examined, while dataset 2 provides the exposure data for the "county CES measure" to be examined, as well as the population subgroups that might be important to stratify by.

# Read in Data files

```{r}
library(tidyverse)
library(janitor)

getwd()

#bring in files and clean up column names
ed <- read.csv('./Project_Data_Sources/scenario_3/chhs_asthma_ed.csv') %>% 
  clean_names()
demog <- read.csv('./Project_Data_Sources/scenario_3/calenviroscreen_scores_demog_2021.csv') %>% 
  clean_names()
ces <- read.csv('./Project_Data_Sources/scenario_3/calenviroscreen_measures_2021.csv') %>% 
  clean_names()

#take a look at columns in each file
head(ed)
head(demog)
head(ces)

#get a sense of numeric variables in each file
summary(ed)
summary(demog)
summary(ces)

#take a look at county names more closely as we need to merge with them and they must be consistent across files
str(ed) 
levels(as.factor(ed$county))
str(demog) #note: county names here are Sentence Case while in ed file they are CAPITALIZED
#Also, in demog, they contain the word "County" while not in ed, and no "CALIFORNIA" in demog
levels(as.factor(demog$county))
str(ces) #here, county is just Capital of county name name, no "County" and variable is named "california_county"
levels(as.factor(ces$california_county))

```

# Examining Data Elements

```{r}
#Identify 5+ data elements required for your specified scenario. Determine the types and the desired type/format for each variable
#Provide basic description of these 5+ elements (summary stats)

#We'll look at: 1) demog$ces_4_0_score, 2) county (demog$county, ed$county, ces$california_county, 3) ed$age_adjusted_ed_visit_rate, 4) ed$year, 5) ces$traffic, 6) ces$pm2_5

#first look at class and type of these variables
class(demog$ces_4_0_score)
typeof(demog$ces_4_0_score)
class(demog$county)
typeof(demog$county)
class(ed$county)
typeof(ed$county)
class(ces$california_county)
typeof(ces$california_county)
class(ed$age_adjusted_ed_visit_rate)
typeof(ed$age_adjusted_ed_visit_rate)
class(ed$year)
typeof(ed$year)
class(ces$traffic)
typeof(ces$traffic)
class(ces$pm2_5)
typeof(ces$pm2_5)
#all measures above look like they are of appropriate type 
#in some cases we may want to use year as categorical, but keep as numeric for now

#now calcualte summary stats for all of the above

#county
levels(as.factor(ed$county))
levels(as.factor(demog$county))
levels(as.factor(ces$california_county))

length(unique(as.factor(ed$county)))
length(unique(as.factor(demog$county)))
length(unique(as.factor(ces$california_county)))

#CES score
summary(demog$ces_4_0_score)

#ED visit rates
summary(ed$age_adjusted_ed_visit_rate)

#year of ED visit - look at as numeric but also as categorical
summary(ed$year)
levels(as.factor(ed$year))

#CES traffic measure (traffic density in vehicle-km per hour per road length)
summary(ces$traffic)

#CES PM 2.5 measure (annual mean PM 2.5)
summary(ces$pm2_5)

```

## Data Cleaning that will be needed

-   County variable names differ between data sets; need to ensure the same case and syntax.

-   Exposures: CES score and environmental measures can be operationalized as continuous but could consider quintiles or quartiles to make them more interpretable.

-   Outcome: ED visit rates can be operationalized as continuous but could also consider categorization to make it more interpretable.

-   Year of ED visit: We need to factor this variable.
